 U.S. National Oceanic and Atmospheric Administration's (NOAA) Analysis 
========================================================

### Reproducible Research: Peer Assessment 1

## Synopsis
Here is going to be cleaned and processed a data base from  U.S. National Oceanic and Atmospheric Administration's (NOAA) ,  the goal is to gain insight about natural  events that occurred from 1950 to 2011, discover which ones cases more injuries and deaths and which ones causes more costs . 

## Data Processing
Load Data  and set working directory. 

```{r}
Data <- read.csv(bzfile("C:/Users/JosePortatil/Documents/GitHub/RepData_PerrAss2/repdata-data-StormData.csv.bz2"))
setwd("C:/Users/JosePortatil/Documents/GitHub/RepData_PerrAss2/")
```

Load library required 

```{r}
require(car)

```

Here the major events are going to ve grouped in some categories because there are too much events an its better have bigger groups with more generics names : 


```{r}
Data$EVTYPE<-with(Data,recode(EVTYPE,"  
                        c('avalanche', 'avalance', 'blizzard', 'chill', 'cold', 'cool', 'glaze', 'hypothermia', 'hyperthermia', 
'ice', 'icy', 'freez', 'frost', 'low temp', 'sleet', 'snow', 'wint')='COLD / ICE';
                        c('below normal precip', 'dry', 'drie', 'drought', 'fire', 'heat', 'high temp', 'hot', 'warm')
='HEAT / DROUGHT / FIRE';
                        c('fog', 'vog')='FOG';
                         c('coast', 'cstl', 'current', 'dam fail', 'dam break', 'drizzle', 'drown', 'erosion', 'erosin', 
'flood', 'floood', 'fld', 'heavy shower', 'high water', 'high waves', 'lake', 'landslump', 'marine', 'precip', 'rain', 
'rising water', 'river', 'rogue wave', 'slide', 'stream', 'sea', 'seiche', 'surf', 'swell', 'tide', 'tidal', 'torrent', 'wet')='RAIN / FLOOD / HIGH SEAS';
                      c('burst', 'cloud', 'depression', 'floyd', 'funnel', 'gust', 'hail', 'hurricane', 'landspout', 
'storm', 'southeast', 'thunderstorm', 'thundertsorm', 'thundestorm', 'tornado', 'torndao', 'tstm', 'turbulence', 'typhoon', 
'wall', 'waterspout', 'water spout', 'wayterspout', 'wind', 'wnd')
='WIND / STORM / TORNADO';
c('lightning', 'ligntning', 'lighting')='LIGHTNING';
c('tsunami', 'volcan')='VOLCANIC ERUPTION / EARTHQUAKE / TSUNAMI';
                        c('dust')='DUST'"))

```


Group by Event type (EVTYPE) and sum injuries in each category. 
```{r}
library(plyr)
INJURIESSummary<-ddply(Data,~EVTYPE,summarise,sum_INJURIES=sum(INJURIES,na.rm=TRUE))
```


Extract events with injuries higher than 100 
```{r}
EventsInjuries<-subset(INJURIESSummary,sum_INJURIES>100)
```

Order Events by INJURIES
```{r}
EventsInjuries <- EventsInjuries[order(EventsInjuries$sum_INJURIES),]
EventsInjuries
```

Group by Event type (EVTYPE) and sum Fatalities in each category.

```{r}
FATALITIESSummary<-ddply(Data,~EVTYPE,summarise,sum_FATALITIES=sum(FATALITIES,na.rm=TRUE))
```


Extract events with fatalities higher than 10
```{r}
EventsFATALITIES<-subset(FATALITIESSummary,sum_FATALITIES>10)
```

Order 
```{r}
EventsFATALITIES <- EventsFATALITIES[order(EventsFATALITIES$sum_FATALITIES),]
EventsFATALITIES
```


Susbet the ones with Billions B (Because the events in Billions of dollars are the ones with biggest costs), then apply GROUP BY , then plot points in billions. 

```{r}
EventsBillions<-subset(Data,PROPDMGEXP=="B")
CostSummary<-ddply(EventsBillions,~EVTYPE,summarise,sum_COST=sum(PROPDMG,na.rm=TRUE))
CostSummary
```

Order By Cost 
```{r}
CostSummary <- CostSummary[order(CostSummary$sum_COST),]
```

## Results

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Injuries: 
```{r ,fig.width=6,fig.height=7}
# Simple Dotplot Injuries 
dotchart(EventsInjuries$sum_INJURIES,labels=EventsInjuries$EVTYPE,cex=.5,
         main="Event type Vs Injuries", 
         xlab="SUm of Injuries caused")
```

Fatalities: 
```{r ,fig.width=6,fig.height=7}
dotchart(EventsFATALITIES$sum_FATALITIES,labels=EventsFATALITIES$EVTYPE,cex=.45,
         main="Event type Vs Fatalities", 
         xlab="SUm of Fatalities caused")
```

 The Tordnado is the most Harmful with the highest Injuries and fatalities followed by excessive heat, flas flood  ... like is showed in this images. 
 
 
2. Across the United States, which types of events have the greatest economic consequences?

```{r ,fig.width=6,fig.height=7}
dotchart(CostSummary$sum_COST,labels=CostSummary $EVTYPE,cex=.5,
         main="Event type Vs Cost [Billions]", 
         xlab="Cost in Billions ")
```

Flood is by far the event woth Greatest economic consequences in Billions of dollars, followed by hurricane and storm surge. 